
\section{Real world graph problems}

\subsection{Scheduling}
\begin{itemize}
\item vertices = jobs that need to be done
\item edges = jobs which require conflicting resources
\end{itemize}
problem: how to decide how many ``periods of work'' needed to complete
all jobs.

Similar problem: table arrangements at a wedding
\begin{itemize}
\item vertices = guests
\item edges = guest that don't get along
\end{itemize}
problem: how many tables?

translation: vertex colorings, chromatic number of a graph
\begin{defn}
A \Xb{vertex coloring} of a graph $G$ is a function $f$ defined on the set of
vertices of $G$. An vertex $n$-coloring is a function $f : V_G \to \{1,
\ldots, n\}$. We say that a vertex coloring is \textbf{proper}\index{vertex
coloring!proper} if $f(v) \neq f(w)$ for adjacent vertices $v, w$.
\end{defn}

\begin{defn}
The chromatic number of a graph, $\chi(G)$, is the minimal number $n$ such
that $G$ has a proper vertex $n$-coloring.
\end{defn}

\subsection{Tournaments}

various teams need to play each other. disjoint pairs of teams can play
simultaneously, but of course the same team can't play at the same time.
How many rounds are needed for teams to play each other?
\begin{itemize}
\item vertices = teams
\item edges = teams who need to play each other
\end{itemize}
problem: how many rounds?

\begin{defn}
An \Xb{edge coloring} of a graph $G$ is a function $f$ defined on the set of
edges of $G$. An edge $n$-coloring is a function $f : E_G \to \{1,
\ldots, n\}$. We say that an edge coloring is \Xb{proper}{proper!edge
coloring} if $f(e) \neq f(e')$ for adjacent vertices $e, e'$.
\end{defn}

\begin{defn}
The \Xb{edge chromatic number} of a graph, $\chi'(G)$, is the minimal
number $n$ such that $G$ has a proper edge $n$-coloring.
\end{defn}

